---
title: "Untitled"
author: "your name here"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```



# 1. Set up

```{r libraries-and-data}
library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # generating model predictions
library(gtsummary) # generating summary tables for models

# abalone data from Hamilton et al. 2022
abalone <- read_xlsx(here("data", "Abalone IMTA_growth and pH.xlsx"))

# Sonadora temperature data from Alonzo, A. 2024
sonadora <- read_csv(here("data", "Temp_SonadoraGradient_Daily.csv"))
```

# 2. Abalone example

Data from Hamilton et al. 2022. "Integrated multi-trophic aquaculture mitigates the effects of ocean acidification: Seaweeds raise system pH and improve growth of juvenile abalone." https://doi.org/10.1016/j.aquaculture.2022.738571

## a. Questions and hypotheses

Question: How does pH predict abalone growth (measured in change in shell surface area per day, mm^-2^ d^-1^)?  

H~0~: pH does not predict abalone growth (change in shell surface area per day, mm^-2^ d^-1^).

H~A~: pH predicts abalone growth (change in shell surface area per day, mm^-2^ d^-1^).

## b. Cleaning

```{r abalone-data-cleaning}
# creating clean data frame
abalone_clean <- abalone |> # start with abalone object
  # clean column names
  clean_names() |> 
  # select columns of interest
  select(mean_p_h, change_in_area_mm2_d_1_25) |> 
  # rename columns
  rename(mean_ph = mean_p_h, 
         change_in_area = change_in_area_mm2_d_1_25) 
```

Don't forget to look at your data! Use `View(abalone_clean)` in the _Console_.

## c. Exploratory data visualization

```{r abalone-scatterplot}
# base layer: ggplot
ggplot(data = abalone_clean,
       aes(x = mean_ph,
           y = change_in_area)) +
  # first layer: points representing abalones
  geom_point(size = 4,
             stroke = 1,
             fill = "firebrick4",
             shape = 21)
```

## d. Abalone model

### Model fitting

```{r abalone-model}
abalone_model <- lm(
  change_in_area ~ mean_ph, # formula: change in area as a function of pH
  data = abalone_clean      # data frame: abalone_clean
)
```

### Diagnostics

```{r abalone-model-diagnostics}

par(mfrow = c(2, 2))   # creating a 2x2 grid
plot(abalone_model)    # plot diagnostic plots

```

### Model summary

```{r abalone-model-summary}

# more information about the model
summary(abalone_model)

```

**insert your notes here**

### Generating predictions

```{r abalone-model-predictions}

# creating a new object called abalone_preds
abalone_preds <- ggpredict(
  abalone_model,      # model object
  terms = "mean_ph"   # predictor (in quotation marks)
)

# display the predictions
abalone_preds

# look at the column names (do this in the Console!!!)
# colnames(abalone_preds)

# look at the actual object (do this in the console!!!)
# View(abalone_preds)


# finding the model prediction at a specific value
ggpredict(
  abalone_model,      # model object
  terms = "mean_ph[8]"   # predictor (in quotation marks) and predictor value in brackets
)

```

**insert your notes here**

### Visualizing model predictions and data

```{r abalone-model-final-figure}

# base layer: ggplot
# using clean data frame
ggplot(data = abalone_clean,
       aes(x = mean_ph,
           y = change_in_area)) +
  # first layer: points representing abalones
  geom_point(size = 4,
             stroke = 1,
             fill = "firebrick4",
             shape = 21) +
  # second layer: ribbon representing confidence interval
  # using predictions data frame
  geom_ribbon(data = abalone_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.1) +
  # third layer: line representing model predictions
  # using predictions data frame
  geom_line(data = abalone_preds,
            aes(x = x,
                y = predicted)) +
  # axis labels
  labs(x = "Mean pH", 
       y = expression("Change in shell area ("*mm^{2}~d^-1*")")) +
  # theme
  theme_minimal()
```

### Creating a table with model coefficients, 95% confidence intervals, and more

**Note:** both these functions are from `gtsummary`.

```{r abalone-table}
tbl_regression(abalone_model,
               # make sure the y-intercept estimate is shown
               intercept = TRUE,
               # changing labels in "Characteristic" column
               label = list(`(Intercept)` = "Intercept",
                            mean_ph = "pH")) |> 
  # changing header text
  modify_header(label = "**Variable**",
                estimate = "**Estimate**") |> 
  # turning table into a flextable (makes things easier to render to word or PDF)
  as_flex_table()
```

# 3. Sonadora temperature example

Data from Ramirez, A. 2024. Sonadora elevational plots: long-term monitoring of air temperature ver 877108. Environmental Data Initiative. https://doi.org/10.6073/pasta/6b66eecae3092d8f2340b5132dec38ab (Accessed 2025-05-14).

## a. Questions and hypotheses

Question: Does elevation (in meters) predict temperature (in °C)?

H~0~: Elevation (m) does not predict temperature (°C).

H~A~: Elevation (m) predicts temperature (°C).

## b. Cleaning and summarizing

```{r temperature-cleaning}
# creating new clean data frame
sonadora_clean <- sonadora |> 
  # clean column names
  clean_names() |> 
  # make the data frame longer
  pivot_longer(cols = plot_250:plot_1000,
               names_to = "plot_name",
               values_to = "temp_c") |> 
  # separate plot name from elevation
  separate_wider_delim(cols = plot_name,
                       delim = "_",
                       names = c("plot", "elevation_m"),
                       cols_remove = FALSE) |> 
  # remove plot column
  select(-plot) |> 
  # make sure elevation is read as a number
  mutate(elevation_m = as.numeric(elevation_m))

# summarizing
sonadora_sum <- sonadora_clean |> 
  # group by plot and elevation
  group_by(plot_name, elevation_m) |> 
  # calculate mean temperature at each elevation
  summarize(mean_temp_c = mean(temp_c, na.rm = TRUE)) |> 
  # undo the group_by function
  ungroup() |> 
  # arrange in order of elevation
  arrange(elevation_m)
```

## c. Exploratory data visualization

```{r temperature-visualization}

# base layer: ggplot
ggplot(data = sonadora_sum, 
       aes(x = elevation_m,
           y = mean_temp_c)) +
  # first layer: points representing temperature at each elevation
  geom_point()

```

## d. Temperature model

```{r temperature-model}

# model
temperature_model <- lm(
  mean_temp_c ~ elevation_m, # formula: response ~ predictor
  data = sonadora_sum # data frame
  )

```

### Diagnostics

```{r temperature-diagnostics}

# diagnostics
par(mfrow = c(2, 2))
plot(temperature_model)

```

### Model summary

```{r temperature-model-summary}

summary(temperature_model)

```

**insert your notes here**

### Generating predictions

```{r temperature-model-predictions}

# model predictions
temperature_preds <- ggpredict(
  temperature_model,      # model object
  terms = "elevation_m"   # predictor (in quotation marks)
)

# calculate the temperature prediction at elevation = 900

```

**insert your notes here**

### Visualizing model predictions

```{r temperature-mode-prediction-visualization}

# base layer
ggplot(data = sonadora_sum, 
       aes(x = elevation_m,
           y = mean_temp_c)) +
  # first layer: temperature at each elevation
  geom_point() +
  # 95% CI ribbon
  # uses model prediction data frame
  geom_ribbon(data = temperature_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2) +
  # model prediction line
  # uses model prediction data frame
  geom_line(data = temperature_preds,
              aes(x = x,
                  y = predicted)) +
  # axis labels
  labs(x = "Elevation (m)",
       y = "Mean temperature (\U00B0 C)") +
  theme_bw()
```

### Creating a table with model coefficients, 95% confidence intervals, and more

```{r temperature-model-table}
tbl_regression(temperature_model,
               # make sure the y-intercept estimate is shown
               intercept = TRUE,
               # changing labels in "Characteristic" column
               label = list(`(Intercept)` = "Intercept",
                            elevation_m = "Elevation (m)")) |> 
  # changing header text
  modify_header(label = "**Variable**",
                estimate = "**Estimate**") |> 
  # turning table into a flextable (makes things easier to render to word or PDF)
  as_flex_table()
```

